\chapter{Introduction}
\label{chap:intro}

\input{conventions}

\section{General Relativistic Hydrodynamics}
\label{sec:gr_hydro}
Fluid dynamics are described by the hydrodynamics equations. In
Euclidean space, for a nonviscous fluid:
\begin{align}
  \label{eqn:hydro_flat_mass}
  \partial_t \rho + \nabla_i S^i       &= 0 \\
  \label{eqn:hydro_flat_mom}
  \partial_t S^j +  \nabla_i T^{ij}    &= f^j \rho \\
  \label{eqn:hydro_flat_ener}
  \partial_t (\frac{1}{2}\rho v^2+u) + \nabla_i(\frac{1}{2}\rho v^2+u+P)v^i &= \rho f_i v^i.
\end{align}
Here $\rho$ is mass density,
$S^i\equiv \rho v^i$ momentum density,
$T^{ij}\equiv \rho v^i v^j + P\delta^{ij}$ the stress tensor,
$f^j=f_j$ the sum of forces per unit mass acting on the fluid,
such as radiation pressure or gravity,
$u$ the specific internal energy, and
$P$ the pressure.
The total specific energy is $e=\rho v^2/2+u$.

This system of five equations (Eqn.~\ref{eqn:hydro_flat_mom} is a composition
of three equations) and six variables $\{\rho,u,P,v^i\}$ is closed by a sixth
equation, the equation of state, relating pressure to internal energy and
density, $P(u,\rho)$.
For an ideal monatomic gas,
we are more familiar with the parametric form of the equation
of state, the two equations $P(\rho,T)=\frac{3}{2m_U}\rho k_{\rm B} T$ and
$u(\rho,T)=\frac{1}{m_U}\rho k_{\rm B} T$.

The five dynamical equations
express for a fluid the conservation of
mass, Eqn.~\ref{eqn:hydro_flat_mass},
momentum, Eqn.~\ref{eqn:hydro_flat_mom}, and
energy, Eqn.~\ref{eqn:hydro_flat_ener}.
As such, they all take the form of hyperbolic conservation equations,
\begin{equation}
  \label{eqn:cons_hyper}
  \partial_t \mathcal{U} + \nabla_i \mathcal{F}^i = \mathcal{S},
\end{equation}
where $\mathcal{U}$ is the conserved quantity,
$\mathcal{F}^i$ its flux, and
$\mathcal{S}$ a source term pumping or draining the conserved quantity.

A hyperbolic equation is fully specified by its initial conditions. In this case
that means choosing an initial configuration $\vec{\mathcal{U}}(t=0)$,
or alternately, specifying the five primitive variables
$\rho(t=0)$, $T(t=0)$, and $v(t=0)$ throughout the physical domain.
Boundary conditions are required to define the behavior of the equations on
the domain boundaries.

A hyperbolic equation like Eqn.~\ref{eqn:cons_hyper}, may be solved via the
method of lines, computing the conservative fields across a spatial domain at
later times, $t>0$ by assuming the spatial dependence at $t=0$ is known. In
one dimension:
\begin{equation}
  \frac{\diff}{\diff t} \mathcal{U} = -\partial_i
\end{equation}

\subsubsection{Relativistic Minkowski}
But in fact, these equations fail when $v$ is close to $c$.
\todo{demonstrate?}
We may write them in their relativistic form:
\begin{align}
  \label{eqn:hydro_sr_mass}
  \nabla_\alpha \rho u^\alpha &= 0 \\
  \label{eqn:hydro_sr_mom}
  \nabla_\alpha T^{\alpha \beta} &= 0.
\end{align}
Here we have introduced $\nabla_\alpha:=(\frac{1}{c}\partial_t,\nabla_i)$,
$u^\alpha:=\gamma(c,v^i)$ the fluid 4-velocity,
$\gamma\equiv(1-v^2/c^2)^{-1/2}$ the Lorentz factor, and
$T^{\alpha\beta}$ the stress-energy tensor:
\begin{eqnarray}
  \label{eqn:se_tensor}
  T^{\alpha\beta} = T_{\rm fluid}^{\alpha\beta} + T_{\rm other}^{\alpha\beta} \\
  T_{\rm fluid}^{\alpha\beta} = (\rho+\frac{u+P}{c^2})u^\alpha u^\beta+P\eta^{\alpha\beta}, \\
\end{eqnarray}
where $\eta^{\alpha\beta}=\eta_{\alpha\beta}:={\rm diag}(-1,1,1,1)$ is the
Minkowski metric, and
$T_{\rm other}^{\alpha\beta}$ is the stress-energy tensor of other fields, like
the radiation field.
Eqns.~\ref{eqn:hydro_sr_mass} and~\ref{eqn:hydro_sr_mom} elegantly express the
conservation of rest mass and momentum-energy.
\todo{what are the conservatives here?}

\section{General Relativistic Radiation Transport}
\label{sec:rad_transport}

